---
title: "ReproRehab Pod: Working With Data in R"
subtitle: "Week 4 — A Short Refresher and an Introduction to Data Manipulation"
author: "Madeline Ratoza"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    number-sections: false
    code-copy: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Today: From Looking at Data to Working With Data

In Week 2, we imported a dataset and performed simple calculations. In Week 3, we explored variable types, loops, and conditionals.

Today, we will use those ideas to answer questions about a new fictional rehab dataset. Most of our time will focus on **data manipulation**: choosing the rows and columns we need, creating new variables, and making small summaries.

You do not need to memorize every function. The goal is to recognize what each tool does and practice adapting an example.

### Learning goals

By the end of this session, you should be able to:

- Explain what a conditional and a loop do.
- Select columns and filter rows in a dataset.
- Sort observations and create a new variable.
- Calculate a summary for the whole dataset or for each group.
- Save your work as a new object while keeping the original data available.

### How to use this tutorial

Read each example, run it in RStudio, and look at the result. Then try the **Your turn** task. Each task includes an expandable hint and answer. Try the hint first; opening an answer is also a useful way to learn.

**Suggested pace:** 5–10 minutes for setup and review, 35–40 minutes for the core activities, and 10–15 minutes for the mini analysis. The missing-data activity is optional.

::: {.callout-note}
## Where do I type?

Create a new R script or Quarto document in your RStudio project. Copy the setup code below, then add your practice code as you work. In a script, run a line or selection with **Command + Enter** on Mac or **Ctrl + Enter** on Windows. In a Quarto document, put R code inside R code chunks.

The webpage shows examples and results. Your own RStudio session is where you run and change the code.
:::

## 1. Meet Our New Fake Dataset

Imagine a small outpatient rehabilitation program wants to describe its participants and their outcomes.

Each row represents **one fictional participant**, with baseline and follow-up recorded in the same row. These are synthetic teaching data, not real patient records.

| Variable | What it represents |
|---|---|
| `participant_id` | A unique fictional ID |
| `age` | Age in years |
| `clinic` | North or South clinic |
| `visits` | Number of visits attended |
| `baseline_score` | Function score at the beginning |
| `followup_score` | Function score at follow-up |
| `completed` | Whether the participant completed the program |

For this exercise, function scores range from 0 to 100, and **higher scores indicate better function**. A positive change therefore represents improvement. These scores and any thresholds we use are invented for practice.

### Create the practice data

Run this entire block once. It creates the same 12 participants for everyone. You do not need to type every value or understand `data.frame()` yet; this is our setup step.

```{r}
clinic_data <- data.frame(
  participant_id = c("P001", "P002", "P003", "P004", "P005", "P006",
                     "P007", "P008", "P009", "P010", "P011", "P012"),
  age = c(34, 67, 52, 45, 71, 29, 63, 48, 76, 39, 58, 65),
  clinic = c("North", "North", "North", "North", "North", "North",
             "South", "South", "South", "South", "South", "South"),
  visits = c(8, 6, 4, 10, 5, 3, 7, 9, 4, 6, 8, 5),
  baseline_score = c(40, 35, 55, 48, 30, 62, 38, 50, 28, 60, 44, 32),
  followup_score = c(55, 43, 58, 66, 40, 61, 50, 65, 30, 69, 57, 38),
  completed = c("Yes", "Yes", "No", "Yes", "Yes", "No",
                "Yes", "Yes", "No", "Yes", "Yes", "Yes")
)
```

Look in your Environment. You should see `clinic_data`.

```{r}
head(clinic_data)
names(clinic_data)
dim(clinic_data)
str(clinic_data)
```

**Pause:** How many participants are there? Which columns contain numbers? Which contain text?

### Your turn: Inspect one variable

Check the variable type of `clinic`.

```{r}
# Write your code here.
```

::: {.callout-tip collapse="true"}
## If you're stuck: Hint

Week 3 used `str()` to inspect types. Use `$` to access one column inside `clinic_data`.
:::

::: {.callout-note collapse="true"}
## Show the answer

```{r}
str(clinic_data$clinic)
```

You should see `chr`, meaning character/text. A variable can represent categories even when R stores its values as text. Today's tools work with these text labels without converting them to factors.
:::

## 2. Quick Refresher: Conditionals and Loops

A **conditional** checks a condition. A **loop** repeats an action.

### A conditional checks one value

Let's check whether the first participant attended at least six visits.

```{r}
first_visits <- clinic_data$visits[1]

if (first_visits >= 6) {
  print("Visit target met")
} else {
  print("Visit target not met")
}
```

`[1]` asks for the first value. `>=` means greater than or equal to. Keep `} else {` together as shown when running the full block.

### A loop repeats an action

```{r}
for (id in clinic_data$participant_id[1:3]) {
  print(id)
}
```

This repeats `print()` for the first three participant IDs. `id` is the name holding the current ID during each pass through the loop.

### Your turn: Apply a condition to every participant

Use `ifelse()` to label each participant as `"Target met"` when they attended **at least six visits**, and `"Below target"` otherwise. Save the result in a new column called `visit_status`.

```{r}
# clinic_data$visit_status <- ifelse(...)
```

::: {.callout-tip collapse="true"}
## If you're stuck: Hint

Use this pattern:

```r
ifelse(condition, "label when TRUE", "label when FALSE")
```

Your condition is `clinic_data$visits >= 6`. Remove the `#` before running a completed line.
:::

::: {.callout-note collapse="true"}
## Show the answer

```{r}
clinic_data$visit_status <- ifelse(
  clinic_data$visits >= 6,
  "Target met",
  "Below target"
)

clinic_data$visit_status
```

This applies the condition across the whole column. Unlike `if`, which expects one TRUE/FALSE value, `ifelse()` can work with many values at once. Both labels describe our practice rule, not a clinical standard.
:::

::: {.callout-note}
## Why aren't we writing more loops?

Loops are useful when we need to repeat a task. Many R functions already operate on whole columns, so we can use them directly for today's calculations and data manipulation.
:::

## 3. Load Our Data Manipulation Tools

We will use **dplyr**, a package for working with tables. It is also part of the tidyverse introduced in Week 2.

If you have not installed it, run this line **once in the Console**:

```{r}
#| eval: false
install.packages("dplyr")
```

Load the package whenever you start a new R session:

```{r}
library(dplyr)
```

Remember: installing puts a package on your computer; `library()` makes it available in the current session.

### Five tools for today

| Function | What it does | Question it helps answer |
|---|---|---|
| `select()` | Keeps columns | Which variables do I need? |
| `filter()` | Keeps rows matching a condition | Which participants do I need? |
| `arrange()` | Sorts rows | In what order do I want to see them? |
| `mutate()` | Adds or changes columns | What new variable do I need? |
| `summarise()` | Calculates summary values | What does this dataset tell me? |

For now, we will put the dataset first inside each function. Later, we will connect a few steps with a pipe.

## 4. Select: Choose Columns

We want a small table with participant IDs, clinics, and visits.

```{r}
visit_table <- select(clinic_data, participant_id, clinic, visits)
visit_table
```

Read this as: take `clinic_data`, keep these three columns, and save the result as `visit_table`.

**Check:** We kept all 12 participants, but fewer columns. `select()` chooses **columns**, not participants.

### Your turn: Make a score table

Create `score_table` containing `participant_id`, `baseline_score`, and `followup_score`. Print it.

```{r}
# Write your code here.
```

::: {.callout-tip collapse="true"}
## If you're stuck: Hint

Start with `score_table <- select(clinic_data, ...)`. Replace `...` with the three column names, separated by commas. Column names inside `select()` do not need quotation marks here.
:::

::: {.callout-note collapse="true"}
## Show the answer

```{r}
score_table <- select(clinic_data, participant_id, baseline_score, followup_score)
score_table
```

You should see 12 rows and three columns. The original `clinic_data` still contains its other columns.
:::

## 5. Filter: Choose Rows

Which participants attended the North clinic?

```{r}
north_data <- filter(clinic_data, clinic == "North")
north_data
```

`==` asks whether values are equal. `"North"` needs quotation marks because it is text. R distinguishes capitalization: `"north"` is different from `"North"`.

| Symbol | Meaning |
|---|---|
| `==` | Equal to |
| `!=` | Not equal to |
| `>` or `<` | Greater than or less than |
| `>=` or `<=` | Greater than or equal to; less than or equal to |
| `&` | AND: both conditions must be true |
| `\|` | OR: at least one condition must be true |

We can combine conditions. This keeps participants from North **who also attended at least six visits**:

```{r}
north_six_visits <- filter(clinic_data, clinic == "North" & visits >= 6)
north_six_visits
```

### Your turn: Who completed the program?

Create `completed_data` containing only participants with `completed == "Yes"`. How many rows remain?

```{r}
# Write your code here.
```

::: {.callout-tip collapse="true"}
## If you're stuck: Hint

Use `filter()` to create the new dataset, then `nrow()` to count its rows. The text value is exactly `"Yes"`.
:::

::: {.callout-note collapse="true"}
## Show the answer

```{r}
completed_data <- filter(clinic_data, completed == "Yes")
nrow(completed_data)
```

There should be **9 participants**. Filtering changes which rows appear in the saved result; it does not delete rows from `clinic_data`.
:::

### Your turn: Combine two conditions

Find participants from the South clinic who are **65 years old or older**. Save them as `south_older`.

```{r}
# Write your code here.
```

::: {.callout-tip collapse="true"}
## If you're stuck: Hint

Use `&` between `clinic == "South"` and `age >= 65`.
:::

::: {.callout-note collapse="true"}
## Show the answer

```{r}
south_older <- filter(clinic_data, clinic == "South" & age >= 65)
south_older
```

You should see **P009 and P012**. The `>=` includes participants who are exactly 65.
:::

## 6. Arrange: Put Rows in Order

Let's sort participants by age, youngest first.

```{r}
age_order <- arrange(clinic_data, age)
age_order
```

`arrange()` sorts from smallest to largest by default. Use `desc()` for largest to smallest:

```{r}
age_order_desc <- arrange(clinic_data, desc(age))
age_order_desc
```

Sorting moves whole rows together. Each participant keeps their own age, scores, and visits.

### Your turn: Most visits first

Create `visits_order` with the participants who attended the most visits at the top.

```{r}
# Write your code here.
```

::: {.callout-tip collapse="true"}
## If you're stuck: Hint

Use `arrange(clinic_data, desc(...))`. Which column goes inside `desc()`?
:::

::: {.callout-note collapse="true"}
## Show the answer

```{r}
visits_order <- arrange(clinic_data, desc(visits))
visits_order
```

**P004** should appear first, with 10 visits.
:::

## 7. Mutate: Create a New Column

How much did each participant's function score change?

```{r}
clinic_working <- mutate(
  clinic_data,
  score_change = followup_score - baseline_score
)

select(clinic_working, participant_id, baseline_score, followup_score, score_change)
```

`mutate()` keeps the existing columns and adds `score_change`. We saved this as `clinic_working`, so our original dataset remains available.

In Week 3, you calculated a change score using a loop. Here, R subtracts the columns across all rows at once.

**Pause:** P001 changed from 40 to 55. What should their change score be? What would a negative change mean?

### Your turn: Add a practice category

In `clinic_working`, add a column called `change_category`. Use `"10 or more"` when `score_change >= 10` and `"Below 10"` otherwise. Save the updated result back to `clinic_working`.

```{r}
# Write your code here.
```

::: {.callout-tip collapse="true"}
## If you're stuck: Hint

Combine `mutate()` with the `ifelse()` you reviewed earlier:

```r
clinic_working <- mutate(
  clinic_working,
  change_category = ifelse(condition, "label when TRUE", "label when FALSE")
)
```

Inside `mutate()`, use column names directly; you do not need `clinic_working$` before them.
:::

::: {.callout-note collapse="true"}
## Show the answer

```{r}
clinic_working <- mutate(
  clinic_working,
  change_category = ifelse(score_change >= 10, "10 or more", "Below 10")
)

select(clinic_working, participant_id, score_change, change_category)
```

Six participants have a change of at least 10. This threshold is only for practicing conditionals; it has no established clinical meaning in these invented data.
:::

## 8. Summarise: Make a Small Summary

What is the average change across all participants?

```{r}
overall_summary <- summarise(
  clinic_working,
  participants = n(),
  average_change = mean(score_change)
)

overall_summary
```

Here, `n()` counts rows and `mean()` calculates an average. `participants` and `average_change` are names we chose for the output columns.

Unlike `mutate()`, which keeps participant-level rows, this `summarise()` produces **one row describing the whole dataset**.

### Your turn: Summarise visits

Create `visit_summary` with the number of participants and their average number of visits.

```{r}
# Write your code here.
```

::: {.callout-tip collapse="true"}
## If you're stuck: Hint

Use `summarise(clinic_working, participants = n(), average_visits = mean(...))`.
:::

::: {.callout-note collapse="true"}
## Show the answer

```{r}
visit_summary <- summarise(
  clinic_working,
  participants = n(),
  average_visits = mean(visits)
)

visit_summary
```

You should see **12 participants** and **6.25 visits** on average.
:::

### A summary for each clinic

`group_by()` tells R to perform the next summary separately for each group.

```{r}
clinic_groups <- group_by(clinic_working, clinic)

clinic_summary <- summarise(
  clinic_groups,
  participants = n(),
  average_change = mean(score_change),
  .groups = "drop"
)

clinic_summary
```

You should get one row for North and one for South, with six participants in each. `.groups = "drop"` leaves the result as an ordinary, ungrouped summary table.

**Think about the result:** Are the averages different? Does a difference in these fictional descriptive summaries establish that one clinic is better? Why might that conclusion require more information?

## 9. Connect Steps With a Pipe

So far, we have saved individual steps as objects. A **pipe**, `|>`, passes the result of one step into the first input of the next function. Read it as **“then.”**

```{r}
completed_scores <- clinic_working |>
  filter(completed == "Yes") |>
  select(participant_id, clinic, score_change) |>
  arrange(desc(score_change))

completed_scores
```

Read this from top to bottom:

1. Start with `clinic_working`.
2. Then keep participants who completed the program.
3. Then keep three columns.
4. Then sort from largest change to smallest.
5. Save the result as `completed_scores`.

The dataset no longer needs to be written inside each function because the pipe supplies it.

::: {.callout-note}
## Which pipe is this?

`|>` is the built-in R pipe and requires R 4.1 or later. You may also see `%>%` in other tutorials. We are using only `|>` today. If your R version is older, use separate steps like those in Sections 4–8, or update R before using these pipe examples.
:::

## 10. Your Turn: A Mini Analysis

The program asks: **Among participants who completed the program, what was the average change at each clinic?**

Use `clinic_working` and complete these steps:

1. Keep participants who completed the program.
2. Group the remaining participants by clinic.
3. Count participants and calculate average change in each clinic.
4. Save the result as `completed_summary` and print it.

You can use a pipe or separate objects. Both approaches are fine.

```{r}
# Write your code here.
```

::: {.callout-tip collapse="true"}
## If you're stuck: Hint 1 — Choose your tools

You need `filter()`, `group_by()`, and `summarise()`, in that order. Your filter condition is `completed == "Yes"`.
:::

::: {.callout-tip collapse="true"}
## If you're stuck: Hint 2 — Start with a pattern

Copy this pattern into RStudio and replace the placeholders before running it:

```r
completed_summary <- clinic_working |>
  filter(____) |>
  group_by(____) |>
  summarise(
    participants = n(),
    average_change = mean(____),
    .groups = "drop"
  )
```
:::

::: {.callout-note collapse="true"}
## Show the answer

```{r}
completed_summary <- clinic_working |>
  filter(completed == "Yes") |>
  group_by(clinic) |>
  summarise(
    participants = n(),
    average_change = mean(score_change),
    .groups = "drop"
  )

completed_summary
```

You should see:

| Clinic | Participants | Average change |
|---|---:|---:|
| North | 4 | 12.75 |
| South | 5 | 11.00 |

If your counts differ, check whether you filtered before summarizing. These summaries describe only program completers; they do not represent every participant.
:::

### Explain your result

In your own words:

- Which participants did you include?
- What does each row in your summary represent?
- Which dataset contains everyone: `clinic_data` or `completed_summary`?
- How would your answer change if you included people who did not complete the program?

## 11. Optional: What Happens When a Value Is Missing?

Real datasets often have missing values, represented by `NA` in R. We will create a separate practice copy with one missing follow-up score.

```{r}
missing_data <- clinic_working
missing_data$followup_score[3] <- NA

missing_data <- mutate(
  missing_data,
  score_change = followup_score - baseline_score
)

mean(missing_data$score_change)
```

The mean returns `NA`: one participant's change cannot be calculated.

If we decide to summarize the available values, we can use `na.rm = TRUE`:

```{r}
mean(missing_data$score_change, na.rm = TRUE)
```

This omits missing values **from this calculation**. It does not fill them in or delete participants from the dataset.

### Your turn: Count available change scores

Create a summary reporting the total number of participants, the number with an available change score, and the average of those available scores.

```{r}
# Write your code here.
```

::: {.callout-tip collapse="true"}
## If you're stuck: Hint

`is.na(score_change)` identifies missing values. `!is.na(score_change)` identifies values that are not missing. `sum()` counts the TRUE values, and `n()` counts all rows.
:::

::: {.callout-note collapse="true"}
## Show the answer

```{r}
missing_summary <- summarise(
  missing_data,
  participants = n(),
  available_scores = sum(!is.na(score_change)),
  average_change = mean(score_change, na.rm = TRUE)
)

missing_summary
```

There are **12 participants**, but only **11 available change scores**. Report how many values contributed to a summary. In a real analysis, investigate why values are missing before deciding how to handle them.
:::

## 12. Save Your Work

Save your script or Quarto document. That code is the recipe for recreating your results.

You can also export a table as a CSV. Run this in your own session when you are ready:

```{r}
#| eval: false
write.csv(completed_summary, "completed_clinic_summary.csv", row.names = FALSE)
```

`row.names = FALSE` prevents an extra row-number column in the exported file. The file is saved in your current working directory; `getwd()` shows that location.

If you want a CSV copy of the original practice dataset for a later session:

```{r}
#| eval: false
write.csv(
  select(clinic_data, participant_id, age, clinic, visits,
         baseline_score, followup_score, completed),
  "week4_clinic_data.csv",
  row.names = FALSE
)
```

That explicitly selects the seven original columns, leaving out the practice column we added in the refresher.

::: {.callout-tip collapse="true"}
## If you're stuck: Common fixes

- **“Could not find function”**: Run `library(dplyr)` for the data manipulation functions.
- **“Object not found”**: Check spelling and run the earlier code that creates the object. Use `names(clinic_data)` to check column names.
- **No rows after filtering**: Check text spelling, capitalization, quotation marks, and your condition.
- **The Console shows `+` instead of `>`**: R is waiting for an unfinished parenthesis, quote, or brace. Press Escape, then run the complete corrected block.
- **Your new column is missing**: Save the result with `<-`. Printing a transformed table alone does not update the saved object.
- **Pipe error**: Check your R version with `R.version.string`. You can use the separate-step examples if `|>` is unavailable.

If results look unexpected, pause and inspect the intermediate table before adding another step.
:::

## Wrap-Up

Today, we used familiar conditions to choose participants, then built new variables and summaries.

Can you explain:

- How `select()` and `filter()` differ?
- Why `==` appears in a filter condition?
- What `desc()` changes?
- How `mutate()` and `summarise()` differ?
- What `group_by()` adds to a summary?
- What the pipe `|>` does?
- Why saving your code matters even after you export a CSV?

**Exit task:** Choose one question about the practice data, write the code to answer it, and describe the result in one sentence. You can reuse and adapt any example from today.
